### 2021-02-25 paper/linkagecluster_bk_paper.Rmd
updated derivations

$$P(\Gamma|\mathbf{Z}, \mathbf{m}, \mathbf{u}, \lambda) =\prod_{j=1}^{n_B}  \left[\lambda^{\mathbf{1}_{z_j < n_A + 1}} (1-\lambda)^{\mathbf{1}_{z_j = n_A + 1}}\prod_{i=1}^{n_A}\mathbf{m}_{ij}^{\mathbf{1}_{z_j = i}}\mathbf{u}_{ij}^{\mathbf{1}_{z_j \neq i}}\right]$$

@@

$$\lambda \sim \text{Beta}(\alpha_{\lambda}, \beta_{\lambda}) $$
The prior for $Z_j$ has equal probability of matching to all records $i\in \mathbf{X}_A$, and non-matching probability governed by $\lambda$. Observe two reasonable choices for $p(Z, \lambda)$:

@@

The $\mathbf{m}$ and $\mathbf{u}$ parameters are updated through standard multinomial-dirichlet mechanics. Thus we have 
$$\mathbf{m^{f}}|\mathbf{Z}, \Gamma \sim \text{Dirichlet}(\alpha^{f1}(\mathbf{Z}), \ldots, \alpha^{fL_f}(\mathbf{Z}))$$
$$\mathbf{u^{f}}|\mathbf{Z}, \Gamma \sim \text{Dirichlet}(\beta^{f1}(\mathbf{Z}), \ldots, \beta^{fL_f}(\mathbf{Z}))$$
where $\alpha_{fl}(\mathbf{Z})= \sum_{i,j} I_{obs}(\gamma_{ij}^f)\mathbf{1}_{\gamma_{ij}^f = l} \mathbf{1}_{z_j = i}$ and $\beta_{fl}(\mathbf{Z})= \sum_{i,j} I_{obs}(\gamma_{ij}^f)\mathbf{1}_{\gamma_{ij}^f = l} \mathbf{1}_{z_j \neq i}$.

@@

@@

>>>>

$$P(\Gamma|\mathbf{Z}, \mathbf{m}, \mathbf{u}, \lambda) =\prod_{j=1}^{n_B}  \prod_{i=1}^{n_A}\mathbf{m}_{ij}^{\mathbf{1}_{z_j = i}}\mathbf{u}_{ij}^{\mathbf{1}_{z_j \neq i}}$$

@@

$$\lambda \sim \text{Beta}(\alpha_{\lambda}, \beta_{\lambda}) $$
The prior for $Z_j$ has equal probability of matching to all records $i\in \mathbf{X}_A$, and non-matching probability governed by $\lambda$. Therefore a  $\lambda \sim \text{Beta}(1, 1)$ corresponds to a prior belief that nonmatches and matches are equally likely, and a $\lambda \sim \text{Beta}(1, \frac{1}{n_A})$ prior corresponds to a uniform prior on the labelling of $\mathbf{Z}$.

@@

We work with the following factorization of the joint distribution:

$$p(\Gamma, \mathbf{Z}, \mathbf{m}, \mathbf{u}, \lambda) = p(\Gamma|\mathbf{Z}, \mathbf{m}, \mathbf{u}) p(\mathbf{Z} | \lambda) p(\mathbf{m}, \mathbf{u}) p(\lambda)$$

This factorization leads to following Gibbs Sampler:

\underline{Sample $\mathbf{m}^{(s+1)}$ $\mathbf{u}^{(s+1)}|\Gamma, \mathbf{Z}^{(s)}$:} The $\mathbf{m}$ and $\mathbf{u}$ parameters are updated through standard multinomial-dirichlet mechanics. Thus we have

$$\mathbf{m^{f}}|\mathbf{Z}, \Gamma \sim \text{Dirichlet}(\alpha^{f1}(\mathbf{Z}), \ldots, \alpha^{fL_f}(\mathbf{Z}))$$
$$\mathbf{u^{f}}|\mathbf{Z}, \Gamma \sim \text{Dirichlet}(\beta^{f1}(\mathbf{Z}), \ldots, \beta^{fL_f}(\mathbf{Z}))$$
where $\alpha_{fl}(\mathbf{Z})= \sum_{i,j} I_{obs}(\gamma_{ij}^f)\mathbf{1}_{\gamma_{ij}^f = l} \mathbf{1}_{z_j = i}$ and $\beta_{fl}(\mathbf{Z})= \sum_{i,j} I_{obs}(\gamma_{ij}^f)\mathbf{1}_{\gamma_{ij}^f = l} \mathbf{1}_{z_j \neq i}$.

\underline{Sample $\lambda^{(s+1)}|\mathbf{Z}^{s}$} As a function of $\lambda$, the linkage structure parameter $\mathbf{Z}$ is sequence of successes and failures, and therefore $p(\mathbf{Z}|\lambda) = \mathcal{L}(\lambda|\mathbf{Z})$  determined only by the number of dupliates $D = \sum_{i=1}^{n_B}\mathbf{1}_{z_j < n_A + 1}$ encoded by $\mathbf{Z}$. Thus we have

$$p(\lambda | \mathbf{Z}) \propto p(\mathbf{Z}|\lambda)p(\lambda)$$
$$\propto \lambda^D (1-\lambda)^{n_B - D} \lambda^{\alpha_{\lambda} -1} (1-\lambda)^{\beta_{\lambda} -1}$$
$$ \propto \lambda^{D + \alpha_{\lambda} - 1} (1-\lambda)^{n_B - D + \beta_{\lambda} -1}$$
$$\implies \lambda^{(s+1)}|\mathbf{Z}^{(s+1)} \sim \text{Beta}(D + \alpha_{\lambda}, n_B - D + \beta_{\lambda})$$

\underline{Sample $\mathbf{Z}^{(s+1)}|\Gamma, \mathbf{m}^{(s+1)}, \mathbf{u}^{(s+1)}, \lambda^{(s+1)}$} Because we sample $Z_j$ independently of all other $Z_{j'}$, we use only the full conditional for an individual $Z_j$. Let $\Gamma_{.j}$ denote the set of $n_A$ comparison vectors with $j \in B$, and note that as a function of $Z_j$, the likelihood $p(\Gamma_{.j}|Z_j, \mathbf{m}, \mathbf{u})$ is a discrete distribution with probalities given by
$$p(\Gamma_{.j}|Z_j = z_j, \mathbf{m}, \mathbf{u}) = \frac{\prod_{i=1}^{n_A}\mathbf{m}_{ij}^{\mathbf{1}_{z_j = i}}\mathbf{u}_{ij}^{\mathbf{1}_{z_j \neq i}}}{\sum_{z_j = 1}^{n_B} \prod_{i=1}^{n_A}\mathbf{m}_{ij}^{\mathbf{1}_{z_j = i}}\mathbf{u}_{ij}^{\mathbf{1}_{z_j \neq i}}} = \frac{\prod_{i=1}^{n_A}\left(\frac{\mathbf{m}_{ij}}{\mathbf{u}_{ij}}\right)^{\mathbf{1}_{z_j = i}}}{\sum_{z_j = 1}^{n_B}\prod_{i=1}^{n_A}\left(\frac{\mathbf{m}_{ij}}{\mathbf{u}_{ij}}\right)^{\mathbf{1}_{z_j = i}}} $$
$$\propto 
\begin{cases} 
    w_{ij}  & z_j \leq n_A; \\
    1 &  z_j  = n_A + 1 \\
\end{cases}$$

where where $w_{ij} = \frac{\mathbf{m}_{ij}}{\mathbf{u}_{ij}} = \frac{P(\boldsymbol{\gamma_{ij}}|Z_j = i)}{P(\boldsymbol{\gamma_{ij}} |Z_j \neq i)}$. The interested reader should note that these are precisely the likelihood ratios used in the Fellegi-Sunter model to classify matches and non-matches, and we therefore refer to $w_{ij}$ as the *Fellegi Sunter weights*.

With the likelihood in this form, we can derive the full conditional
$$p(Z_j|\Gamma_{.j}, \mathbf{m} ,\mathbf{u}, \lambda) \propto p(\Gamma_{.j}| Z_j, \mathbf{m} ,\mathbf{u}) P(Z_j|\lambda)$$

$$\propto \left(\sum_{i=1}^{n_A}w_{ij}\mathbf{1}_{z_j = i} + \mathbf{1}_{z_j = n_A + 1}\right)\left(\lambda\sum_{i=1}^{n_A}\frac{1}{n_A}\mathbf{1}_{z_j = i} + (1-\lambda)\mathbf{1}_{z_j = n_A + 1}\right)$$
$$= \frac{\lambda}{n_A}\sum_{i=1}^{n_A}w_{ij}\mathbf{1}_{z_j = i} + (1-\lambda)\mathbf{1}_{z_j = n_A + 1} $$
$$ \implies Z_j^{(s+1)} | \mathbf{m}, \mathbf{u}, \Gamma, \lambda \propto
\begin{cases} 
    \frac{\lambda}{n_A}w_{ij}   & z_j \leq n_A; \\
     1-\lambda &  z_j  = n_A + 1 \\
\end{cases}$$

@@

\newpage

@@

\newpage 
Observe two reasonable choices for $p(Z, \lambda)$:

The $\mathbf{m}$ and $\mathbf{u}$ parameters are updated through standard multinomial-dirichlet mechanics. Thus we have

$$\mathbf{m^{f}}|\mathbf{Z}, \Gamma \sim \text{Dirichlet}(\alpha^{f1}(\mathbf{Z}), \ldots, \alpha^{fL_f}(\mathbf{Z}))$$
$$\mathbf{u^{f}}|\mathbf{Z}, \Gamma \sim \text{Dirichlet}(\beta^{f1}(\mathbf{Z}), \ldots, \beta^{fL_f}(\mathbf{Z}))$$
where $\alpha_{fl}(\mathbf{Z})= \sum_{i,j} I_{obs}(\gamma_{ij}^f)\mathbf{1}_{\gamma_{ij}^f = l} \mathbf{1}_{z_j = i}$ and $\beta_{fl}(\mathbf{Z})= \sum_{i,j} I_{obs}(\gamma_{ij}^f)\mathbf{1}_{\gamma_{ij}^f = l} \mathbf{1}_{z_j \neq i}$.

### 2021-02-25 paper/linkagecluster_bk_paper.tex
updated derivations

\[P(\Gamma|\mathbf{Z}, \mathbf{m}, \mathbf{u}, \lambda) =\prod_{j=1}^{n_B}  \left[\lambda^{\mathbf{1}_{z_j < n_A + 1}} (1-\lambda)^{\mathbf{1}_{z_j = n_A + 1}}\prod_{i=1}^{n_A}\mathbf{m}_{ij}^{\mathbf{1}_{z_j = i}}\mathbf{u}_{ij}^{\mathbf{1}_{z_j \neq i}}\right]\]

@@

\[\lambda \sim \text{Beta}(\alpha_{\lambda}, \beta_{\lambda}) \] The
prior for \(Z_j\) has equal probability of matching to all records
\(i\in \mathbf{X}_A\), and non-matching probability governed by
\(\lambda\). Observe two reasonable choices for \(p(Z, \lambda)\):

@@

The \(\mathbf{m}\) and \(\mathbf{u}\) parameters are updated through
standard multinomial-dirichlet mechanics. Thus we have
\[\mathbf{m^{f}}|\mathbf{Z}, \Gamma \sim \text{Dirichlet}(\alpha^{f1}(\mathbf{Z}), \ldots, \alpha^{fL_f}(\mathbf{Z}))\]
\[\mathbf{u^{f}}|\mathbf{Z}, \Gamma \sim \text{Dirichlet}(\beta^{f1}(\mathbf{Z}), \ldots, \beta^{fL_f}(\mathbf{Z}))\]
where
\(\alpha_{fl}(\mathbf{Z})= \sum_{i,j} I_{obs}(\gamma_{ij}^f)\mathbf{1}_{\gamma_{ij}^f = l} \mathbf{1}_{z_j = i}\)
and
\(\beta_{fl}(\mathbf{Z})= \sum_{i,j} I_{obs}(\gamma_{ij}^f)\mathbf{1}_{\gamma_{ij}^f = l} \mathbf{1}_{z_j \neq i}\).

@@

(THIS WOULD GO IN APPENDIX) Our new model specification becomes

@@

>>>>

\[P(\Gamma|\mathbf{Z}, \mathbf{m}, \mathbf{u}, \lambda) =\prod_{j=1}^{n_B}  \prod_{i=1}^{n_A}\mathbf{m}_{ij}^{\mathbf{1}_{z_j = i}}\mathbf{u}_{ij}^{\mathbf{1}_{z_j \neq i}}\]

@@

\[\lambda \sim \text{Beta}(\alpha_{\lambda}, \beta_{\lambda}) \] The
prior for \(Z_j\) has equal probability of matching to all records
\(i\in \mathbf{X}_A\), and non-matching probability governed by
\(\lambda\). Therefore a \(\lambda \sim \text{Beta}(1, 1)\) corresponds
to a prior belief that nonmatches and matches are equally likely, and a
\(\lambda \sim \text{Beta}(1, \frac{1}{n_A})\) prior corresponds to a
uniform prior on the labelling of \(\mathbf{Z}\).

@@

We work with the following factorization of the joint distribution:

\[p(\Gamma, \mathbf{Z}, \mathbf{m}, \mathbf{u}, \lambda) = p(\Gamma|\mathbf{Z}, \mathbf{m}, \mathbf{u}) p(\mathbf{Z} | \lambda) p(\mathbf{m}, \mathbf{u}) p(\lambda)\]

This factorization leads to following Gibbs Sampler:

\underline{Sample $\mathbf{m}^{(s+1)}$ $\mathbf{u}^{(s+1)}|\Gamma, \mathbf{Z}^{(s)}$:}
The \(\mathbf{m}\) and \(\mathbf{u}\) parameters are updated through
standard multinomial-dirichlet mechanics. Thus we have

\[\mathbf{m^{f}}|\mathbf{Z}, \Gamma \sim \text{Dirichlet}(\alpha^{f1}(\mathbf{Z}), \ldots, \alpha^{fL_f}(\mathbf{Z}))\]
\[\mathbf{u^{f}}|\mathbf{Z}, \Gamma \sim \text{Dirichlet}(\beta^{f1}(\mathbf{Z}), \ldots, \beta^{fL_f}(\mathbf{Z}))\]
where
\(\alpha_{fl}(\mathbf{Z})= \sum_{i,j} I_{obs}(\gamma_{ij}^f)\mathbf{1}_{\gamma_{ij}^f = l} \mathbf{1}_{z_j = i}\)
and
\(\beta_{fl}(\mathbf{Z})= \sum_{i,j} I_{obs}(\gamma_{ij}^f)\mathbf{1}_{\gamma_{ij}^f = l} \mathbf{1}_{z_j \neq i}\).

\underline{Sample $\lambda^{(s+1)}|\mathbf{Z}^{s}$} As a function of
\(\lambda\), the linkage structure parameter \(\mathbf{Z}\) is sequence
of successes and failures, and therefore
\(p(\mathbf{Z}|\lambda) = \mathcal{L}(\lambda|\mathbf{Z})\) determined
only by the number of dupliates
\(D = \sum_{i=1}^{n_B}\mathbf{1}_{z_j < n_A + 1}\) encoded by
\(\mathbf{Z}\). Thus we have

\[p(\lambda | \mathbf{Z}) \propto p(\mathbf{Z}|\lambda)p(\lambda)\]
\[\propto \lambda^D (1-\lambda)^{n_B - D} \lambda^{\alpha_{\lambda} -1} (1-\lambda)^{\beta_{\lambda} -1}\]
\[ \propto \lambda^{D + \alpha_{\lambda} - 1} (1-\lambda)^{n_B - D + \beta_{\lambda} -1}\]
\[\implies \lambda^{(s+1)}|\mathbf{Z}^{(s+1)} \sim \text{Beta}(D + \alpha_{\lambda}, n_B - D + \beta_{\lambda})\]

\underline{Sample $\mathbf{Z}^{(s+1)}|\Gamma, \mathbf{m}^{(s+1)}, \mathbf{u}^{(s+1)}, \lambda^{(s+1)}$}
Because we sample \(Z_j\) independently of all other \(Z_{j'}\), we use
only the full conditional for an individual \(Z_j\). Let \(\Gamma_{.j}\)
denote the set of \(n_A\) comparison vectors with \(j \in B\), and note
that as a function of \(Z_j\), the likelihood
\(p(\Gamma_{.j}|Z_j, \mathbf{m}, \mathbf{u})\) is a discrete
distribution with probalities given by
\[p(\Gamma_{.j}|Z_j = z_j, \mathbf{m}, \mathbf{u}) = \frac{\prod_{i=1}^{n_A}\mathbf{m}_{ij}^{\mathbf{1}_{z_j = i}}\mathbf{u}_{ij}^{\mathbf{1}_{z_j \neq i}}}{\sum_{z_j = 1}^{n_B} \prod_{i=1}^{n_A}\mathbf{m}_{ij}^{\mathbf{1}_{z_j = i}}\mathbf{u}_{ij}^{\mathbf{1}_{z_j \neq i}}} = \frac{\prod_{i=1}^{n_A}\left(\frac{\mathbf{m}_{ij}}{\mathbf{u}_{ij}}\right)^{\mathbf{1}_{z_j = i}}}{\sum_{z_j = 1}^{n_B}\prod_{i=1}^{n_A}\left(\frac{\mathbf{m}_{ij}}{\mathbf{u}_{ij}}\right)^{\mathbf{1}_{z_j = i}}} \]
\[\propto 
\begin{cases} 
    w_{ij}  & z_j \leq n_A; \\
    1 &  z_j  = n_A + 1 \\
\end{cases}\]

where where
\(w_{ij} = \frac{\mathbf{m}_{ij}}{\mathbf{u}_{ij}} = \frac{P(\boldsymbol{\gamma_{ij}}|Z_j = i)}{P(\boldsymbol{\gamma_{ij}} |Z_j \neq i)}\).
The interested reader should note that these are precisely the
likelihood ratios used in the Fellegi-Sunter model to classify matches
and non-matches, and we therefore refer to \(w_{ij}\) as the
\emph{Fellegi Sunter weights}.

With the likelihood in this form, we can derive the full conditional
\[p(Z_j|\Gamma_{.j}, \mathbf{m} ,\mathbf{u}, \lambda) \propto p(\Gamma_{.j}| Z_j, \mathbf{m} ,\mathbf{u}) P(Z_j|\lambda)\]

\[\propto \left(\sum_{i=1}^{n_A}w_{ij}\mathbf{1}_{z_j = i} + \mathbf{1}_{z_j = n_A + 1}\right)\left(\lambda\sum_{i=1}^{n_A}\frac{1}{n_A}\mathbf{1}_{z_j = i} + (1-\lambda)\mathbf{1}_{z_j = n_A + 1}\right)\]
\[= \frac{\lambda}{n_A}\sum_{i=1}^{n_A}w_{ij}\mathbf{1}_{z_j = i} + (1-\lambda)\mathbf{1}_{z_j = n_A + 1} \]
\[ \implies Z_j^{(s+1)} | \mathbf{m}, \mathbf{u}, \Gamma, \lambda \propto
\begin{cases} 
    \frac{\lambda}{n_A}w_{ij}   & z_j \leq n_A; \\
     1-\lambda &  z_j  = n_A + 1 \\
\end{cases}\]

@@

Our new model specification becomes

@@

\newpage

Observe two reasonable choices for \(p(Z, \lambda)\):

The \(\mathbf{m}\) and \(\mathbf{u}\) parameters are updated through
standard multinomial-dirichlet mechanics. Thus we have

\[\mathbf{m^{f}}|\mathbf{Z}, \Gamma \sim \text{Dirichlet}(\alpha^{f1}(\mathbf{Z}), \ldots, \alpha^{fL_f}(\mathbf{Z}))\]
\[\mathbf{u^{f}}|\mathbf{Z}, \Gamma \sim \text{Dirichlet}(\beta^{f1}(\mathbf{Z}), \ldots, \beta^{fL_f}(\mathbf{Z}))\]
where
\(\alpha_{fl}(\mathbf{Z})= \sum_{i,j} I_{obs}(\gamma_{ij}^f)\mathbf{1}_{\gamma_{ij}^f = l} \mathbf{1}_{z_j = i}\)
and
\(\beta_{fl}(\mathbf{Z})= \sum_{i,j} I_{obs}(\gamma_{ij}^f)\mathbf{1}_{\gamma_{ij}^f = l} \mathbf{1}_{z_j \neq i}\).
